\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-05-31 2-cubble.Rmd
work on subsection 2.3 functionalities

@@

## Functions and methods in cubble

*Print all the methods with methods(class = "cubble_df")*

### dplyr methods

@@

The created \code{cubble} is a subclass of the \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested \code{cubble}. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

```{r echo = TRUE}
cb_nested |> mutate(rain_day = sum(ts$prcp != 0))
```

### cubble specific methods

The three cubble specific functions are \code{face_temporal()}, \code{face_spatial()}, and \code{unfold}.

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and stores the spatial variables as an attribute. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the \code{cubble} object just created into its long form:

```{r echo = TRUE}
(cb_long <- cb_nested |> face_temporal())
```

```{r face, echo = FALSE, fig.align="center", out.width = "100%", fig.cap = "An illustration of function \\code{face\\_temporal} and \\code{face\\_spatial}. In the first row, \\code{face\\_temporal} switches a \\code{cubble} object from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \\code{face\\_spatial} switches a \\code{cubble} object back to the nested form from the long form and shifts focus back to the spatial aspect."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.002.png"))
```

Wrangling spatio-temporal data can be seen as an iterative process in the spatial and temporal dimensions. Switching the focus back to the site-variable face can be accomplished by the function \code{face_spatial()}, which is the inverse of \code{face_temporal()}. The second row of Figure \ref{fig:face} illustrates the function, which is used as follows:

```{r echo = TRUE}
(cb_back <- cb_long |> face_spatial())
#identical(cb_nested, cb_back)
```

>>>>

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and stores the spatial variables as an attribute. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure  illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face.

@@

## Functionalities

Table \@ref(tab:funs) summarises functions implemented in the \pkg{cubble} package.

Table: (\#tab:funs) Functions implemented in the \pkg{cubble} package

|Category | Details |  Functions|
|---|---|--------------|
|base R | subset | `[`, `[<-`, `names<-`, `[[`, `[[<-`| 
|tidyverse | dplyr |`dplyr_row_slice`, `dplyr_col_modify`, `dplyr_reconstruct`, `select`, `mutate`, `arrange`, `filter`, `summarise`, `slice`, `group_by`, `ungroup`|
|tidyverse| tidyr| | 
|tidyverse| pillar| `type_sum` | 
|sf| | | 
|tsibble| | | 
|cubble | basics| `face_temporal`, `face_spatial`, `unfold`|
| | attributes| `key`, `key_vars`, `key_data`, `index`, `index_var`, `coords`, `spatial`|
| |hierarchy|  `switch_key`|
| |matching |   `match_sites`, `match_spatial`, `match_temporal`| 
| |glyph map | `geom_glyph`, `geom_glyph_box`, `geom_glyph_line`|

### dplyr and tidyr

@@

The toy example below demonstrates using dplyr verbs on a nested cubble: 
```{r echo = TRUE}
cb_nested %>% 
  mutate(avg_temp = mean(ts$tmax, na.rm = TRUE)) %>% 
  select(-elev) %>% 
  filter(avg_temp > 26) %>% 
  arrange(-avg_temp)
```

### cubble

The three basic cubble functions, \code{face_temporal()}, \code{face_spatial()}, and \code{unfold()}, are introduced in this section before more advanced functionalities in Section \@ref(others).

The pair of verbs, \code{face_temporal()} and \code{face_spatial()}, pivot the cubble object between the spatial and temporal face of the multivariate spatio-temporal cube, as illustrated in Figure \ref{fig:face}. The code below uses the function \code{face_temporal()} and \code{face_spatial()} to transform between a nested cubble (\code{cb_nested}) and a long one (\code{cb_long}), introduced earlier in subsection \@ref(object):

```{r echo = TRUE}
identical(face_temporal(cb_nested), cb_long)
identical(face_spatial(cb_long), cb_nested)
```

The pair of verbs are exact inverse and apply both functions on a cubble object will result in the object itself:

```{r echo = TRUE}
identical(face_spatial(face_temporal(cb_nested)), cb_nested)
identical(face_temporal(face_spatial(cb_long)), cb_long)
```

```{r face, echo = FALSE, fig.align="center", out.width = "100%", fig.cap = "An illustration of the function \\code{face\\_temporal()} and \\code{face\\_spatial()}: \\code{face\\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \\code{face\\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.002.png"))
```

### 2023-05-31 jss/main.tex
work on subsection 2.3 functionalities

\hypertarget{functions-and-methods-in-cubble}{%
\subsection{Functions and methods in cubble}\label{functions-and-methods-in-cubble}}

\emph{Print all the methods with methods(class = ``cubble\_df'')}

\hypertarget{dplyr-methods}{%
\subsubsection{dplyr methods}\label{dplyr-methods}}

@@

The created \code{cubble} is a subclass of the \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested \code{cubble}. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

\begin{CodeChunk}
\begin{CodeInput}
R> cb_nested |> mutate(rain_day = sum(ts$prcp != 0))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form
# extent:   [144.8321, -37.98, 145.0964, -37.6655]
# temporal: prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name          wmo_id rain_day ts      
  <chr>       <dbl> <dbl> <dbl> <chr>          <dbl>    <int> <list>  
1 ASN00086038  145. -37.7  78.4 essendon air~  95866        3 <tibble>
2 ASN00086077  145. -38.0  12.1 moorabbin ai~  94870        3 <tibble>
3 ASN00086282  145. -37.7 113.  melbourne ai~  94866        3 <tibble>
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{cubble-specific-methods}{%
\subsubsection{cubble specific methods}\label{cubble-specific-methods}}

The three cubble specific functions are \code{face_temporal()}, \code{face_spatial()}, and \code{unfold}.

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and stores the spatial variables as an attribute. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the \code{cubble} object just created into its long form:

\begin{CodeChunk}
\begin{CodeInput}
R> (cb_long <- cb_nested |> face_temporal())
\end{CodeInput}
\begin{CodeOutput}
# cubble:  key: id [3], index: date, long form
# extent:  2020-01-01 -- 2020-01-10 [1D], no gaps
# spatial: long [dbl], lat [dbl], elev [dbl], name [chr], wmo_id
#   [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00086038 2020-01-01     0  26.8  11  
2 ASN00086038 2020-01-02     0  26.3  12.2
3 ASN00086038 2020-01-03     0  34.5  12.7
4 ASN00086038 2020-01-04     0  29.3  18.8
5 ASN00086038 2020-01-05    18  16.1  12.5
# i 25 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial}. In the first row, \code{face\_temporal} switches a \code{cubble} object from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a \code{cubble} object back to the nested form from the long form and shifts focus back to the spatial aspect.}\label{fig:face}
\end{figure}
\end{CodeChunk}

Wrangling spatio-temporal data can be seen as an iterative process in the spatial and temporal dimensions. Switching the focus back to the site-variable face can be accomplished by the function \code{face_spatial()}, which is the inverse of \code{face_temporal()}. The second row of Figure \ref{fig:face} illustrates the function, which is used as follows:

\begin{CodeChunk}
\begin{CodeInput}
R> (cb_back <- cb_long |> face_spatial())
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form
# extent:   [144.8321, -37.98, 145.0964, -37.6655]
# temporal: prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name              wmo_id ts      
  <chr>       <dbl> <dbl> <dbl> <chr>              <dbl> <list>  
1 ASN00086038  145. -37.7  78.4 essendon airport   95866 <tibble>
2 ASN00086077  145. -38.0  12.1 moorabbin airport  94870 <tibble>
3 ASN00086282  145. -37.7 113.  melbourne airport  94866 <tibble>
\end{CodeOutput}
\begin{CodeInput}
R> #identical(cb_nested, cb_back)
\end{CodeInput}
\end{CodeChunk}

>>>>

\hypertarget{functionalities}{%
\subsection{Functionalities}\label{functionalities}}

Table \ref{tab:funs} summarises functions implemented in the \pkg{cubble} package.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7000}}@{}}
\caption{\label{tab:funs} Functions implemented in the \pkg{cubble} package}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Details
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Details
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endhead
base R & subset & \texttt{{[}}, \texttt{{[}\textless{}-}, \texttt{names\textless{}-}, \texttt{{[}{[}}, \texttt{{[}{[}\textless{}-} \\
tidyverse & dplyr & \texttt{dplyr\_row\_slice}, \texttt{dplyr\_col\_modify}, \texttt{dplyr\_reconstruct}, \texttt{select}, \texttt{mutate}, \texttt{arrange}, \texttt{filter}, \texttt{summarise}, \texttt{slice}, \texttt{group\_by}, \texttt{ungroup} \\
tidyverse & tidyr & \\
tidyverse & pillar & \texttt{type\_sum} \\
sf & & \\
tsibble & & \\
cubble & basics & \texttt{face\_temporal}, \texttt{face\_spatial}, \texttt{unfold} \\
& attributes & \texttt{key}, \texttt{key\_vars}, \texttt{key\_data}, \texttt{index}, \texttt{index\_var}, \texttt{coords}, \texttt{spatial} \\
& hierarchy & \texttt{switch\_key} \\
& matching & \texttt{match\_sites}, \texttt{match\_spatial}, \texttt{match\_temporal} \\
& glyph map & \texttt{geom\_glyph}, \texttt{geom\_glyph\_box}, \texttt{geom\_glyph\_line} \\
\bottomrule()
\end{longtable}

\hypertarget{dplyr-and-tidyr}{%
\subsubsection{dplyr and tidyr}\label{dplyr-and-tidyr}}

@@

The toy example below demonstrates using dplyr verbs on a nested cubble:

\begin{CodeChunk}
\begin{CodeInput}
R> cb_nested %>% 
+   mutate(avg_temp = mean(ts$tmax, na.rm = TRUE)) %>% 
+   select(-elev) %>% 
+   filter(avg_temp > 26) %>% 
+   arrange(-avg_temp)
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [2], index: date, nested form
# extent:   [144.8321, -37.7276, 144.9066, -37.6655]
# temporal: prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat name              wmo_id avg_temp ts      
  <chr>       <dbl> <dbl> <chr>              <dbl>    <dbl> <list>  
1 ASN00086282  145. -37.7 melbourne airport  94866     26.6 <tibble>
2 ASN00086038  145. -37.7 essendon airport   95866     26.5 <tibble>
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{cubble-1}{%
\subsubsection{cubble}\label{cubble-1}}

The three basic cubble functions, \code{face_temporal()}, \code{face_spatial()}, and \code{unfold()}, are introduced in this section before more advanced functionalities in Section \ref{others}.

The pair of verbs, \code{face_temporal()} and \code{face_spatial()}, pivot the cubble object between the spatial and temporal face of the multivariate spatio-temporal cube, as illustrated in Figure \ref{fig:face}. The code below uses the function \code{face_temporal()} and \code{face_spatial()} to transform between a nested cubble (\code{cb_nested}) and a long one (\code{cb_long}), introduced earlier in subsection \ref{object}:

\begin{CodeChunk}
\begin{CodeInput}
R> identical(face_temporal(cb_nested), cb_long)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> identical(face_spatial(cb_long), cb_nested)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

The pair of verbs are exact inverse and apply both functions on a cubble object will result in the object itself:

\begin{CodeChunk}
\begin{CodeInput}
R> identical(face_spatial(face_temporal(cb_nested)), cb_nested)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> identical(face_temporal(face_spatial(cb_long)), cb_long)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

@@

\caption{An illustration of the function \code{face\_temporal()} and \code{face\_spatial()}: \code{face\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \code{face\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables.}\label{fig:face}
\end{figure}
\end{CodeChunk}
